Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-03-09 ThesisJP/Background/background.tex
ULtrasound imaging.

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}), which has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512). The newly introduced extensions, AVX and AVX2, adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point operations can be performed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32, 32 bit, floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS},
\end{equation}
hence a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz. This adds
\begin{equation}
16*40*1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

@@

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

@@

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space. This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}. A cardiac probe is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide. This kind of image, where A-mode images are acquired from multiple spatial locations in the $xz$-plane, is known as a B-mode image (B for brightness). If A-mode images are steered in the $yz$-plane as well we get what is known as a volumetric or 3D ultrasound image.

If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m[n]$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array \textit{weight vector} or \textit{apodization}, and $z[n]$ is the output A-mode image in the direction given by $\Delta_m[n]$.  Delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}. If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays have to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays therefore do not vary within an A-mode image.

When generating a B-mode image we have to insure that the spatial sampling is sufficient. We have already seen that the lateral resolution of an ultrasound imaging probe is given by (\ref{eq:res}).  Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support is governed by the sum of the transmit (\nom{tx}{Transmit}) and receive (\nom{rx}{Receive}) aperture. The lateral resolution of an active imaging system is then found to be
\begin{align}
\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
\end{align}
\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}In order to properly sample an imaging sector of $\theta$ radians we have to acquire 
\begin{align}
N_\theta = \frac{\theta}{\delta\theta}
\end{align}
\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}A-mode images. If \nom{$r$}{Imaging range.} is the maximum range we want to image we get an imaging frame rate of
\begin{align}
FR = \frac{c}{2rN_\theta}
\end{align} 
frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe from the earlier example we then get a frame rate of $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short impulse consist of many frequency components. A short ultrasound pulse, which would yield high axial resolution, therefore requires a large frequency bandwidth. A ultrasound probe has normally a relative bandwidth between 50 and 100 \%. This corresponds to transmitting a pulse which are between 2 and 1 oscillation long. In the front-end of an ultrasound imaging system a high sampling frequency is used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. Number of samples in one A-mode image is then found to be:
\begin{align}
N = \frac{2r}{c}B_wf_c.
\end{align}
\nomi{$N$}{Number of samples in range}where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency} is the center frequency of the transmitted pulse. The total number of samples in an B-mode image is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
For our cardiac probe we have to process close to $50000$ $M$-long vectors of element data every $1/55$ second.

Time v.s. phase delays.

Key selling points of ultrasound imaging: cost, safety and real-time user interaction.

@@

Small section about heart anatomy (Fig.\,\ref{fig:human_heart}).

\subsection{Accelerators in medical ultrasound imaging}
Ultrasound imaging systems generates an incredible amount of data. A modern system capable of 3D imaging will generate several tera bytes of data in the front end of the system during a clinical scan. The data rate from a 2D scan is "only" in order giga bytes per second. To handle these amount of data, ultrasound imaging systems have relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline. However, as the computationally power of CPUs and GPUs have increased, ASICs, e.g. for scan conversion, have gradually been converted into software \cite{Guracar2013}. One of  the last remaining ASICs is the beamformer, whose job is to steer and focus the ultrasound beam in space. This operation involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe. The input data rate is therefor very high. Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave ultrasound beamforming algorithm \cite{Tanter2014}. Beamforming is therefore finally moving into software to. The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}. Software beamforming opens up new possibilities for advanced processing. One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), another is advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.

>>>>

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}), which has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512). The newly introduced extensions, AVX and AVX2, adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point operations can be performed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32, 32 bit, floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS},
\end{equation}
hence a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz. This adds
\begin{equation}
16*40*1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

@@

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

@@

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}. A cardiac probe is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide. This kind of image, where A-mode images are acquired from multiple spatial locations in the $xz$-plane, is known as a B-mode image (B for brightness). If A-mode images are steered in the $yz$-plane as well we get what is known as a volumetric B-mode or 3D ultrasound image.

If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array \textit{weight vector} or \textit{apodization}, and $z[n]$ is the output A-mode image in the direction given by $\Delta_m[n]$.  Delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}. If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays has to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays therefore do not vary within an A-mode image.

When generating a B-mode image we have to insure that the spatial and temporal sampling is sufficient. We have already seen that the lateral resolution of an ultrasound imaging probe is given by (\ref{eq:res}).  Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support is governed by the sum of the transmit (\nom{tx}{Transmit}) and receive (\nom{rx}{Receive}) aperture. The lateral resolution of an active imaging system is then found to be
\begin{align}
\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
\end{align}
\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}Then, in order to properly sample an imaging sector of $\theta$ radians we have to acquire 
\begin{align}
N_\theta = \frac{\theta}{\delta\theta}
\end{align}
\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}A-mode images uniformly distributed across the sector. If \nom{$r$}{Imaging range.} is the maximum range we want to image we get an imaging frame rate of
\begin{align}
FR = \frac{c}{2rN_\theta}
\end{align} 
frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe from the earlier example we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short impulse consist of many frequency components. A short ultrasound pulse, which would yield high axial resolution, therefore requires a large frequency bandwidth. A ultrasound probe has normally a relative bandwidth between 50 and 100 \%. This corresponds to transmitting a pulse which are between 2 and 1 oscillation long. In the front-end of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{\nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. Number of samples in one A-mode image is then found to be:
\begin{align}
N = \frac{2r}{c}B.
\end{align}
\nomi{$N$}{Number of samples in range}where \nom{$B$}{Frequency bandwidth} is the bandwidth, $B = B_wf_c$ (where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency of ultrasound pulse} is the center frequency of the transmitted pulse). The total number of samples in an B-mode image is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
For our cardiac probe has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of element data every $1/55$ second.

%Time v.s. phase delays. It is well described in the paper using phase delays.

@@

Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation. There are no gamma rays or high intensity magnetic fields involved, just sound pressure below well controlled limits. It is also far less costly than other imaging modalities and offers real-time inspection of the human body.  Ultrasound imaging is well know for its use in fetal imaging during pregnancy, however, in this thesis we will focus on different modality, namely cardiac ultrasound imaging. An illustration of the human heart is therefore included in Fig.\,\ref{fig:human_heart}. Throughout the thesis we will refer to some of the parts shown in Fig.\,\ref{fig:human_heart}. Especially note the four chambers of the heart, the left and right atrium, and the left and right ventricle. The tissue separating the left and right ventricle is know as the inter-ventricular septum. The left side is the strongest and hardest working part of the heart, and it is responsible of pumping blood out to the whole body through the aortic valve. When blood flows out the aorta it is important that the mitral valve, that separates the left atrium and ventricle, is closed and does not leak. The left ventricle and the mitral valve therefore receives special attention during cardiac ultrasound examinations. When visualizing 3D ultrasound images we will have a strong focus on visibility of cardiac tissue, and note for later that the endocardium refers to the innermost layer of the cardiac muscle.

\subsection{GPUs in medical ultrasound imaging}
Ultrasound imaging systems generates an incredible amount of data. A modern system capable of 3D imaging will generate several tera bytes of data in the front end of the system during a clinical scan. The data rate from a 2D scan is "only" in order giga bytes per second. To handle these amount of data, ultrasound imaging systems have relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline. However, as the computationally power of CPUs and GPUs have increased, ASICs, e.g. for converting data from a polar to a Cartesian grid (known as scan conversion), have gradually been converted into software \cite{Guracar2013}. One of  the last remaining ASICs is the beamformer, whose job is to steer and focus the ultrasound beam in space. This operation involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe. The input data rate is therefor very high. Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}. Beamforming is therefore finally moving into software too. The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}. Software beamforming opens up new possibilities for advanced processing of ultrasound data. One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}). The increased computational power introduced by GPUs makes it also possible to include advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.
